\documentclass{idc_msc}

\title{Advanced Topics in IP Networks \\\large Lecture 13}
\date{2019-01-10 \\ Last edited \currenttime\ \today}
\author{Lecture by Dr. Anat Bremler-Barr\\Typeset by Steven Karas}

\AtEndDocument{\vfill\bibliographystyle{plain}\bibliography{../biblio}{}}

\begin{document}

\maketitle

\paragraph{Disclaimer}

These notes are based on the lectures for the course Advanced Topics in IP Networks, taught by Dr. Anat Bremler-Barr at IDC Herzliyah in the fall semester of 2018/2019.
Sections may be based on the lecture slides prepared by Dr. Anat Bremler-Barr.

\nocite{Varghese:2004:NAI:1203994}
\nocite{Crovella:2006:IMI:1196480}
\nocite{Kurose:2002:CNT:549735}

\section{Agenda}

\begin{itemize}
  \item Projects
\end{itemize}

\section{Homework}

The p4 project is due on Sunday.

On that topic, the current direction is that there are dozens of startups that sell p4 hardware, but the large companies have not yet started.

\section{Project}

\subsection{Requirements}

\begin{itemize}
  \item A github repo with the project code
  \item A report detailing the project. In the format of a conference paper (2-4 pages).
    \begin{enumerate}
      \item Introduction
      \item Related Work
      \item Technical Details
      \item Evaluation/Discussion
      \item Conclusion
      \item Bibliography - minimum of two citations - one for the paper, and one for the github repo
    \end{enumerate}
  \item Video - record the project running and show the results
  \item Presentation - all members must present. 20 minutes.
  As a matter of course, we can reuse sections from the original slides with proper attribution.
\end{itemize}

\subsection{Timeline}

On January 17th, project registration will open. At most two groups are allowed to work on a proposed project.
Outside proposals are welcome, but should come from the list of approved conferences.

No later than January 31st, the project proposal should be submitted.
The proposal should be no longer than two pages, and include technical details on how to implement the project.
No grade will be given on the proposal.

No later than February 16th, a final decision will be made regarding the proposal.

No later than April 1st, projects must be submitted and presentation scheduled.

Each of these stages may be done earlier with no benefit or penalty.

\section{Fin}

The rest of the lecture covered the various project proposals.

Have a fun winter break!

\end{document}
